---
title: "An introduction to CellSPA"
author:
- name: Yingxin Lin
  affiliation: School of Mathematics and Statistics, The University of Sydney, Australia
output:
  BiocStyle::html_document:
    toc_float: true
package: BiocStyle
vignette: >
  %\VignetteIndexEntry{CellSPA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r global-options, include=FALSE}
knitr::opts_chunk$set(
    message=FALSE,
    warning=FALSE,
    collapse=TRUE,
    comment="#>"
)
library(BiocStyle)
```




```{r, include=FALSE}

library(CellSPA)
library(ggplot2)
library(ggthemes)
library(scater)
library(SingleCellExperiment)
library(SpatialExperiment)
#Why do I need to call these in the rmd,the dependent functions fail if I dont..?
library(sf)
library(spatstat)
library(lwgeom)
theme_set(theme_bw())
```


# Read BIDCell output

```{r, cache=TRUE}
source("C:\\Users\\dmech\\OneDrive - The University of Sydney (Students)\\PhD\\Metrics\\CellSPA\\R\\cellshape_metrics.R")
# data_dir <- system.file("extdata/BIDCell_cells_breast1/", package = "CellSPA")
# tiff_path <- system.file("extdata/BIDCell_cells_breast1.tif", package = "CellSPA")
data_dir <- "data/BIDCell_cells_breast1"
tiff_path <- "data/BidCell_cells_breast1.tif"
spe <- readBIDCell(data_dir,
                   tiff_path = tiff_path,
                   method_name = "BIDCell")
```


```{r, cache=TRUE}
spe <- processingSPE(spe,
                     qc_range = list(total_transciprts = c(20, 2000),
                                     total_genes = c(20, Inf)))

```


```{r}
#subset a set of cells for illustration
spe <- CellSPA::subset(spe, 1:50)

```


# Baseline metrics


```{r}

spe <- generatePolygon(spe)
spe <- calBaselineAllMetrics(spe, verbose = TRUE)
head(rowData(spe))
head(colData(spe))
```



# Expression similarity

## Processing reference

```{r}
sce_ref_full <- readRDS(system.file("extdata/sce_FFPE_full.rds", 
                            package = "CellSPA"))
sce_ref <- processingRef(sce_ref_full, 
                         celltype = sce_ref_full$graph_cluster_anno, 
                         subset_row = rownames(spe))
sce_ref
```


## Calculating expression similarity 

```{r}
spe <- calExpressionCorrelation(spe, 
                                sce_ref,
                                ref_celltype = sce_ref$celltype,
                                method = c("pearson", "cosine"),
                                spe_exprs_values = "logcounts",
                                ref_exprs_values = "mean")
head(colData(spe))
```

```{r}
spe <- calExpressionCorrelation(spe, 
                                sce_ref,
                                ref_celltype = sce_ref$celltype,
                                method = c("pearson", "cosine"),
                                spe_exprs_values = "logcounts",
                                ref_exprs_values = "prop_detected")
head(colData(spe))
```

```{r}
spe <- calAggExpressionCorrelation(spe,
                                   celltype = "mean_celltype_correlation",
                                   sce_ref = sce_ref,
                                   ref_celltype = "celltype",
                                   method = c("pearson"),
                                   spe_exprs_values = "logcounts",
                                   ref_exprs_values = "mean")
diag(spe@metadata$CellSPA$similarity_metrics$agg_mean_correlation)
```
```{r}
spe <- calAggExpressionCorrelation(spe,
                                   celltype = "mean_celltype_correlation",
                                   sce_ref = sce_ref,
                                   ref_celltype = "celltype",
                                   method = c("pearson"),
                                   spe_exprs_values = "logcounts",
                                   ref_exprs_values = "prop_detected")
diag(spe@metadata$CellSPA$similarity_metrics$agg_prop_detected_correlation)
```



## Calculate marker F1 purity


```{r}
# Generate a positive marker list based on the reference data
positive_marker_list <- generateMarkerList(sce_ref, type = "positive")
positive_marker_list
negative_marker_list <- generateMarkerList(sce_ref, type = "negative", t = 1)
negative_marker_list
```




```{r}
spe <- calMarkerPurity(spe,
                       celltype = "mean_celltype_correlation",
                       marker_list = positive_marker_list,
                       marker_list_name = "positive")
head(colData(spe))
```



```{r}
spe <- calMarkerPurity(spe,
                       celltype = "mean_celltype_correlation",
                       marker_list = negative_marker_list,
                       marker_list_name = "negative")
head(colData(spe))
```


## Calculate marker expressed pct


```{r}

spe <- calMarkerPct(spe,
                    celltype = "mean_celltype_correlation",
                    marker_list = positive_marker_list,
                    marker_list_name = "positive")
head(colData(spe))
```



```{r}

spe <- calMarkerPct(spe,
                    celltype = "mean_celltype_correlation",
                    marker_list = negative_marker_list,
                    marker_list_name = "negative")
head(colData(spe))
```



# Spatial Variation

```{r}
spe <- calSpatialMetricsDiversity(spe, 
                                  celltype = "mean_celltype_correlation")
```


```{r}
df <- spe@metadata$CellSPA$spatialMetricsDiversity$results
ggplot(df, aes(x = x_bin, y = y_bin, fill = `cellTypeProp_ECM1+ Malignant`)) +
    geom_tile() +
    scale_fill_viridis_c() +
    theme(aspect.ratio = 1, 
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank()) +
    labs(fill = "ECM1+ Malignant %") 


ggplot(df, aes(x = entropy, y = cv_total_transciprts, 
               color = `cellTypeProp_ECM1+ Malignant`)) +
    geom_point() +
    scale_color_viridis_c() +
    theme(aspect.ratio = 1, 
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank()) +
    labs(color = "ECM1+ Malignant %")
```


# Neighbour purity

```{r}
nn_celltype_pair <- c("B Cells", "CD4 T|CD8 T")
neg_markers <- list("B Cells" = c("CD3C", "CD3E", "CD8A"),
                    "CD4 T|CD8 T" = c("MS4A1", "CD79A", "CD79B"))
spe <- calNegMarkerVsDist(spe,
                          "mean_celltype_correlation",
                          nn_celltype_pair,
                          neg_markers)
spe@metadata$CellSPA$`negMarkerExprs_vs_dist`
```


# Read 10x output


```{r}
tenX_output_dir <- system.file("extdata/10x_output_subset", package = "CellSPA")
tenX_output_tif <- system.file("extdata/10x_output_subset/10x_from_csv_subset.tif", package = "CellSPA")
spe_10x <- readXenium(tenX_output_dir,
                      tiff_path = tenX_output_tif)


```


```{r}

keep_idx <- spatialCoords(spe_10x)[, 2] <= max(spatialCoords(spe)[, 1]) &
    spatialCoords(spe_10x)[, 2] >= min(spatialCoords(spe)[, 1])
spe_10x <- CellSPA::subset(spe_10x, which(keep_idx))

spe_10x <- processingSPE(spe_10x,
                         qc_range = list(total_transciprts = c(20, 2000),
                                         total_genes = c(20, Inf)))
spe_10x <- CellSPA::subset(spe_10x, 1:500)
spe_10x
```





Run CellSPA with a wrapper function `CellSPA`.

```{r}
spe_10x <- CellSPA(spe_10x,
                   spe_celltype = NULL,
                   sce_ref = sce_ref,
                   ref_celltype = sce_ref$celltype,
                   positive_marker_list = positive_marker_list,
                   negative_marker_list = negative_marker_list,
                   nn_celltype_pair = nn_celltype_pair,
                   nn_neg_markers_list = neg_markers,
                   exprs_values = "logcounts",
                   use_BPPARAM = BiocParallel::SerialParam(),
                   verbose = TRUE)
```


# Visualisation

## Within method

We can use `plotColData` in `scater` package to visualise the baseline metrics.

```{r}
scater::plotColData(spe, "total_transciprts", 
                    x = "mean_celltype_correlation", 
                    colour_by = "mean_celltype_correlation") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


scater::plotColData(spe, "elongation", 
                    x = "mean_celltype_correlation", 
                    colour_by = "mean_celltype_correlation") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```


```{r}
spe_list <- list(BIDCell = spe,
                 `10x` = spe_10x)
```



# Session Info

```{r}
sessionInfo()
```


